---
title: "Fitting infectious disease models to data in R"
subtitle: "Practical 2: Bayesian inference and Markov chain Monte Carlo"
date: "2026-09-14"
author: "James Mba Azam, PhD"
format:
  html:
    theme:
      light: flatly
      dark: darkly
execute:
  echo: true
  warning: false
  message: false
  cache: true
engine: knitr
editor:
  markdown:
    wrap: 72
---

::: {.callout-tip collapse="true" icon="false"}
### How to work with this Quarto document

This document is created using **Quarto**, a powerful publishing system
that combines text, code, and output. Here are some tips for working
with it:

#### Rendering the document:

- **In RStudio/Positron**: Click the "Render" button or use
  `Ctrl+Shift+K` (Windows/Linux) or `Cmd+Shift+K` (Mac)
- **Command line**: Run `quarto render filename.qmd` in your terminal
- **Alternative formats**: Change `format: html` to `format: pdf` in the
  YAML header to render as PDF

#### Running code interactively:

- **Individual lines**: Place cursor on a line and press `Ctrl+Enter`
  (Windows/Linux) or `Cmd+Enter` (Mac)
- **Code chunks**: Click the green "play" button in the top-right corner
  of any code chunk, or use `Ctrl+Shift+Enter` (Windows/Linux) or
  `Cmd+Shift+Enter` (Mac)
- **All chunks**: Use "Run All" from the Run menu or `Ctrl+Alt+R`
  (Windows/Linux) or `Cmd+Option+R` (Mac)

#### Best practices:

- Run code chunks sequentially from top to bottom on your first pass
- Experiment with parameter values by modifying the code and re-running
  chunks
- Save your work frequently using `Ctrl+S` (Windows/Linux) or `Cmd+S`
  (Mac)
:::

## Learning Objectives

By the end of this tutorial, you will be able to:

- Explain the roles of the prior, the likelihood and the posterior in
  Bayesian inference
- Write down a prior for the parameters of an SIR model and check it by
  simulating from it
- Explain why the posterior cannot usually be computed directly, and how
  Monte Carlo sampling gets around this
- Implement the Metropolis algorithm from scratch and apply it first to
  a known distribution, then to the posterior of the SIR model fitted to
  outbreak data
- Tune a proposal distribution and judge from trace plots, acceptance
  rates, $\hat{R}$ and effective sample sizes whether a chain can be
  trusted
- Summarise a posterior sample: point estimates, credible intervals,
  derived quantities such as $R_0$, and predictions with uncertainty

## Prerequisites

This tutorial follows on from the [least squares and maximum likelihood
practical](01_least_squares_mle_practicals.qmd), and assumes:

- Basic R programming (data manipulation, plotting, writing functions)
- Familiarity with the SIR model and with the Poisson likelihood used in
  the maximum likelihood section of the previous practical
- Elementary probability (probability distributions, conditional
  probability)

**Required R packages:**

```{r setup}
# Load required packages
library(deSolve)   # solving the SIR differential equations
library(ggplot2)   # plotting
library(MASS)      # mvrnorm(): multivariate normal draws for the proposal

set.seed(1978)     # so that your chains match the ones in this document
```

## Introduction

In the previous practical we fitted an SIR model to the 1978 influenza
outbreak in a British boarding school by least squares and by maximum
likelihood. Both methods returned a single *best* value for each
parameter, with uncertainty (standard errors, profile likelihood
intervals) obtained as a second step.

Bayesian inference asks a different question. Instead of *"which
parameter values fit best?"* it asks *"how plausible is each possible
parameter value, given the data?"*. The answer is a whole probability
distribution over the parameters, the **posterior distribution**, and
every quantity we report (estimates, intervals, predictions) is read off
that distribution.

The price is computational: for a model like ours the posterior cannot
be written down in closed form. **Markov chain Monte Carlo (MCMC)** is
the family of algorithms that lets us draw samples from it anyway. In
this tutorial you will build the simplest MCMC algorithm, the
**Metropolis algorithm**, yourself, so that you understand what packages
such as Stan, NIMBLE and `brms` are doing for you.

### The Data: 1978 Influenza Outbreak

We use the same data as before: daily counts of pupils confined to bed
during a 14-day influenza outbreak in a boarding school of 763 pupils.

```{r}
# Load and prepare the data
flu <- read.csv("./data/influenza_england_1978_school.csv")
flu$day <- seq_len(nrow(flu))    # days 1 to 14
flu$cases <- flu$in_bed          # pupils in bed with flu on each day

ggplot(flu, aes(x = day, y = cases)) +
  geom_line() +
  geom_point() +
  labs(x = "Day", y = "Number of Cases",
       title = "Influenza Outbreak at British Boarding School, 1978",
       subtitle = "Daily number of students in bed with flu") +
  theme_minimal()
```

### The Model

The SIR model, and the function that returns the model's prediction of
the number of pupils in bed on days 1 to 14, are the same as in the
maximum likelihood section of the previous practical. We start the
epidemic at day 0 with one infectious pupil.

```{r}
# SIR model for deSolve
sir_model <- function(t, y, params) {
  with(as.list(c(y, params)), {
    dS <- -beta * S * I
    dI <- beta * S * I - gamma * I
    dR <- gamma * I
    list(c(dS, dI, dR))
  })
}

init <- c(S = 762, I = 1, R = 0)   # one index case in a school of 763
times <- 0:14                       # day 0 is the index case, data are days 1-14

# Model prediction of pupils in bed on days 1 to 14
predict_cases <- function(beta, gamma) {
  out <- ode(y = init, times = times, func = sir_model,
             parms = c(beta = beta, gamma = gamma))
  out[-1, "I"]
}

# A quick check: the trajectory for beta = 0.002, gamma = 0.5
round(predict_cases(beta = 0.002, gamma = 0.5))
```

## Bayesian inference: prior, likelihood, posterior

### Bayes' rule

Write $\theta$ for the parameters (here $\beta$ and $\gamma$) and $y$
for the data. Bayes' rule says

$$p(\theta \mid y) = \frac{p(y \mid \theta)\, p(\theta)}{p(y)}$$

- $p(\theta)$ is the **prior**: what we believe about the parameters
  *before* seeing the data
- $p(y \mid \theta)$ is the **likelihood**: the probability of the data
  for given parameter values. This is exactly the quantity we maximised
  in the previous practical
- $p(\theta \mid y)$ is the **posterior**: what we believe *after*
  seeing the data
- $p(y)$ is a normalising constant that makes the posterior integrate to
  one. It does not depend on $\theta$

Because $p(y)$ is the same for every $\theta$, the shape of the
posterior is determined by the product of the other two:

$$\text{posterior} \propto \text{likelihood} \times \text{prior}$$

A useful way to think about this (Kruschke & Liddell, 2018): before the
data, credibility is spread over the possible parameter values according
to the prior. The data then *reallocate* credibility towards the values
that explain them well and away from those that do not. Bayes' rule is
the bookkeeping.

### Parameter transformation

As in the maximum likelihood practical, we work with the logarithms of
the rates,

$$b = \log \beta, \qquad g = \log \gamma,$$

so that the parameters we sample can take any real value while $\beta$
and $\gamma$ themselves stay positive. Throughout, `theta` is the vector
`c(b, g)`.

### The prior

The prior is part of the model, so it has to be stated. We use normal
priors on the log-rates, which is the same as lognormal priors on the
rates themselves:

- $b \sim N(\log 0.0013,\ 1)$. The centre comes from
  $\beta = R_0 \gamma / N$ with $R_0 = 2$, $\gamma = 0.5$ and $N = 763$,
  i.e. a fairly ordinary influenza outbreak. A standard deviation of 1
  on the log scale means "within a factor of $e \approx 2.7$ either
  way", which is deliberately vague.
- $g \sim N(\log 0.5,\ 1)$: an infectious period of around two days,
  again with a wide spread.

```{r}
log_prior <- function(theta) {
  dnorm(theta[1], mean = log(0.0013), sd = 1, log = TRUE) +
    dnorm(theta[2], mean = log(0.5), sd = 1, log = TRUE)
}
```

We work with **log** densities throughout: products of many small
probabilities become sums, and nothing underflows to zero.

### Check the prior by simulating from it

A prior on $\beta$ is hard to have intuition about, but an epidemic
curve is not. So draw parameter values from the prior, run the model,
and look at the epidemics it produces. This is a **prior predictive
check**.

```{r}
prior_curves <- do.call(rbind, lapply(1:50, function(k) {
  b <- rnorm(1, mean = log(0.0013), sd = 1)
  g <- rnorm(1, mean = log(0.5), sd = 1)
  data.frame(sim = k, day = 1:14, cases = predict_cases(exp(b), exp(g)))
}))

ggplot() +
  geom_line(data = prior_curves, aes(x = day, y = cases, group = sim),
            colour = "steelblue", alpha = 0.4) +
  geom_point(data = flu, aes(x = day, y = cases), size = 2) +
  labs(x = "Day", y = "Pupils in bed",
       title = "50 epidemics simulated from the prior") +
  theme_minimal()
```

The prior allows anything from no outbreak at all to the whole school in
bed within three days. That is what a vague prior should look like: it
covers the data comfortably without being centred on them.

::: callout-tip
### Exercise: a more informative prior

Change the prior on $g$ to $N(\log 0.5,\ 0.2)$ (an infectious period of
roughly 1.5–3 days) and re-run the prior predictive check. What changes?
What would you need to know about influenza to justify such a prior?
:::

### The likelihood

As in the previous practical, we treat each day's count as a Poisson
draw around the model's prediction:

$$y_i \sim \text{Poisson}(\lambda_i), \qquad \lambda_i = I(t_i;\ \beta, \gamma)$$

```{r}
log_likelihood <- function(theta) {
  predicted <- predict_cases(beta = exp(theta[1]), gamma = exp(theta[2]))
  # The ODE solver can return tiny negative numbers once the epidemic has
  # burnt out; treat any such trajectory as impossible
  if (any(!is.finite(predicted)) || any(predicted < 0)) return(-Inf)
  sum(dpois(flu$cases, lambda = predicted, log = TRUE))
}
```

### The posterior, up to a constant

```{r}
log_posterior <- function(theta) log_prior(theta) + log_likelihood(theta)

# Evaluate it at a few points
log_posterior(c(log(0.002), log(0.5)))
log_posterior(c(log(0.0022), log(0.48)))
log_posterior(c(log(0.001), log(0.5)))
```

Notice what we can and cannot do. We can **evaluate** the (unnormalised)
log posterior at any $\theta$ we like: it is one call to the ODE solver.
What we cannot do is compute the normalising constant $p(y)$, which
would require integrating over every combination of $\beta$ and
$\gamma$, or draw samples from the posterior directly.

### Looking at the posterior on a grid

With only two parameters we can get a picture of the posterior by
evaluating it on a grid.

```{r}
beta_grid <- seq(0.0020, 0.0025, length.out = 40)
gamma_grid <- seq(0.40, 0.56, length.out = 40)

grid <- expand.grid(beta = beta_grid, gamma = gamma_grid)
grid$log_post <- mapply(function(b, g) log_posterior(c(log(b), log(g))),
                        grid$beta, grid$gamma)
grid$post <- exp(grid$log_post - max(grid$log_post))   # relative to the peak

ggplot(grid, aes(x = beta, y = gamma, fill = post)) +
  geom_raster() +
  geom_contour(aes(z = post), colour = "grey30", breaks = c(0.01, 0.1, 0.5, 0.9)) +
  scale_fill_viridis_c(name = "relative\nposterior", option = "inferno", direction = -1) +
  labs(x = expression(beta), y = expression(gamma),
       title = "Posterior density on a 40 x 40 grid (1,600 model runs)") +
  theme_minimal()
```

Two things to notice:

1.  The posterior is concentrated in a small region. With 14 days of
    data on hundreds of pupils, the likelihood dominates our vague
    prior.
2.  The contours are tilted: higher $\beta$ goes with higher $\gamma$.
    The parameters are **correlated** in the posterior, because a faster
    epidemic with a shorter infectious period produces a similar curve
    to a slower one with a longer infectious period. We will need to
    keep this in mind when we tune our sampler.

A grid is a fine way to *look at* a two-parameter posterior. It is not a
way to fit models: 15 values per parameter is 225 model runs for two
parameters, 50,625 for four, and over 11 million for six. We need a
method whose cost does not explode with the number of parameters.

## Monte Carlo and the Metropolis algorithm

### Learning about a distribution from samples

If we had a large sample of draws from the posterior, we could estimate
anything we wanted about it: the mean and median of each parameter, a
95% interval from the 2.5% and 97.5% quantiles, the distribution of a
derived quantity like $R_0 = \beta N / \gamma$ by simply computing it
for every draw. This is the **Monte Carlo** idea, and it is how we will
report our results.

The problem is that we can only *evaluate* the posterior, up to a
constant. **MCMC** solves exactly this problem: it generates a sequence
(a *chain*) of parameter values in which each new value depends only on
the previous one, arranged so that, in the long run, the chain visits
each region of parameter space in proportion to the posterior density.

### The Metropolis algorithm

The Metropolis algorithm (Metropolis et al., 1953) is the simplest MCMC
method:

1.  Start at some value $\theta^{(0)}$.
2.  **Propose** a new value by adding random noise to the current one:
    $\theta^* = \theta + \varepsilon$, with $\varepsilon$ drawn from a
    symmetric *proposal distribution*, e.g. $N(0, \sigma^2)$.
3.  Compute the ratio of posterior densities
    $$r = \frac{p(\theta^* \mid y)}{p(\theta \mid y)}.$$ The unknown
    constant $p(y)$ appears in numerator and denominator and cancels:
    **we never need it**.
4.  Draw $u \sim \text{Uniform}(0, 1)$. If $u < r$, **accept**: the
    chain moves to $\theta^*$. Otherwise **reject**: the chain stays at
    $\theta$.
5.  Record the current value of $\theta$ (the old value again, if the
    proposal was rejected) and go back to step 2.

Step 4 means that a proposal with *higher* posterior density is always
accepted ($r > 1$), while a proposal with *lower* density is accepted
with probability $r$. The chain therefore drifts towards regions of high
posterior density but can also move downhill, which is what lets it
explore the whole distribution rather than just climb to the mode. It
can be shown (Chib & Greenberg, 1995) that this acceptance rule makes
the posterior the *stationary distribution* of the chain: once the chain
has forgotten where it started, its values are draws from the posterior.

### Implementing it for one parameter

Here is the whole algorithm in R. It takes any function that returns a
log density, so we can test it on a distribution whose answer we know.

```{r}
metropolis_1d <- function(log_target, n_iter, init, proposal_sd) {
  samples <- numeric(n_iter)
  theta <- init
  log_p <- log_target(theta)
  n_accepted <- 0
  for (i in seq_len(n_iter)) {
    # 1. propose a move: current value plus symmetric noise
    proposal <- theta + rnorm(1, mean = 0, sd = proposal_sd)
    log_p_proposal <- log_target(proposal)
    # 2. accept with probability min(1, p(proposal) / p(theta)); on the log
    #    scale that is: accept if log(u) < log_p_proposal - log_p
    if (log(runif(1)) < log_p_proposal - log_p) {
      theta <- proposal
      log_p <- log_p_proposal
      n_accepted <- n_accepted + 1
    }
    # 3. record where the chain is (the old value again, if we rejected)
    samples[i] <- theta
  }
  list(samples = samples, acceptance_rate = n_accepted / n_iter)
}
```

### A test on a known target

MCMC does not know anything about Bayesian inference: it samples from
whatever distribution you give it a log density for. So let's point it
at a $N(5, 2)$ distribution, where we know what the samples should look
like.

```{r}
log_target <- function(theta) dnorm(theta, mean = 5, sd = 2, log = TRUE)

run <- metropolis_1d(log_target, n_iter = 5000, init = 0, proposal_sd = 5)

run$acceptance_rate
mean(run$samples)   # should be close to 5
sd(run$samples)     # should be close to 2
```

```{r}
trace <- data.frame(iteration = seq_along(run$samples), theta = run$samples)

ggplot(trace, aes(x = iteration, y = theta)) +
  geom_line(colour = "grey30") +
  geom_hline(yintercept = 5, colour = "tomato", linewidth = 1) +
  labs(title = "Trace plot: where the chain has been") +
  theme_minimal()

ggplot(trace, aes(x = theta)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "grey80") +
  stat_function(fun = dnorm, args = list(mean = 5, sd = 2), colour = "tomato", linewidth = 1) +
  labs(title = "Sampled distribution against the true N(5, 2) density") +
  theme_minimal()
```

The **trace plot** shows the chain bouncing around the true mean with a
constant spread, and the histogram of the samples matches the true
density. A well-behaved trace is often described as a "hairy
caterpillar": no trend, the same width all the way along.

### Tuning the proposal

The proposal standard deviation is a **tuning parameter**: it is not
part of the model, but it decides how well the sampler works.

```{r}
tuning <- do.call(rbind, lapply(c(0.2, 5, 100), function(sd) {
  r <- metropolis_1d(log_target, n_iter = 1000, init = 5, proposal_sd = sd)
  data.frame(proposal_sd = sprintf("sd = %g (acceptance %.0f%%)", sd, 100 * r$acceptance_rate),
             iteration = 1:1000, theta = r$samples)
}))

ggplot(tuning, aes(x = iteration, y = theta)) +
  geom_line(colour = "grey30") +
  geom_hline(yintercept = 5, colour = "tomato") +
  facet_wrap(~ proposal_sd, ncol = 1) +
  theme_minimal()
```

- With a **narrow** proposal almost every move is accepted, but each
  move is tiny: the chain wanders slowly and takes a long time to cover
  the distribution.
- With a **wide** proposal most proposals land far out in the tails and
  are rejected: the chain sits still for long stretches.
- In between, the chain mixes well. For random-walk proposals of this
  kind, the theory (Roberts, Gelman & Gilks, 1997) says to aim for an
  acceptance rate of about 44% for one parameter, falling to about 23%
  as the number of parameters grows.

::: callout-tip
### Exercise: proposal width

Try proposal standard deviations of 1, 2 and 10. Which gives the
acceptance rate closest to 44%? Does the histogram of samples still look
right for the badly tuned chains if you run them for long enough (say
50,000 iterations)? What does that tell you about the difference between
*correctness* and *efficiency*?
:::

### Burn-in

The starting value is arbitrary, and the first part of the chain is
spent travelling from it to the region of high density. Those early
samples are not draws from the target and must be discarded: the
**burn-in**.

```{r}
burn <- do.call(rbind, lapply(c(5, 40, 100), function(start) {
  r <- metropolis_1d(log_target, n_iter = 500, init = start, proposal_sd = 5)
  data.frame(start = paste("start at", start), iteration = 1:500, theta = r$samples)
}))

ggplot(burn, aes(x = iteration, y = theta)) +
  annotate("rect", xmin = 0, xmax = 100, ymin = -Inf, ymax = Inf, alpha = 0.15, fill = "steelblue") +
  geom_line(colour = "grey30") +
  geom_hline(yintercept = 5, colour = "tomato") +
  facet_wrap(~ start) +
  labs(title = "Chains from three starting values; shaded region = a 100-iteration burn-in") +
  theme_minimal()
```

How much to discard is decided *after* looking at the chains. Being
generous costs little, as long as enough samples remain. A good starting
value (for instance the maximum likelihood estimate) shortens the
burn-in.

## Fitting the SIR model by MCMC

### A sampler for several parameters

For the SIR model $\theta = (b, g)$ has two components. The only change
to the algorithm is that the proposal is now a two-dimensional normal
draw, described by a covariance matrix rather than a single standard
deviation. `MASS::mvrnorm()` does the drawing.

```{r}
metropolis <- function(log_target, n_iter, init, proposal_cov) {
  samples <- matrix(NA, nrow = n_iter, ncol = length(init))
  theta <- init
  log_p <- log_target(theta)
  if (!is.finite(log_p)) stop("the starting value has zero posterior density")
  n_accepted <- 0
  for (i in seq_len(n_iter)) {
    proposal <- theta + MASS::mvrnorm(1, mu = rep(0, length(init)),
                                      Sigma = proposal_cov)
    log_p_proposal <- log_target(proposal)
    if (log(runif(1)) < log_p_proposal - log_p) {
      theta <- proposal
      log_p <- log_p_proposal
      n_accepted <- n_accepted + 1
    }
    samples[i, ] <- theta
  }
  list(samples = samples, acceptance_rate = n_accepted / n_iter)
}
```

### A pilot run

We do not yet know how wide the posterior is, so we start with a rough,
uncorrelated proposal (standard deviation 0.1 on the log scale for each
parameter, i.e. moves of about 10%) and a starting value at the centre
of the prior. This run takes a few seconds: every iteration solves the
SIR equations once.

```{r}
pilot <- metropolis(log_posterior, n_iter = 3000,
                    init = c(log(0.0013), log(0.5)),
                    proposal_cov = diag(c(0.1, 0.1)^2))
pilot$acceptance_rate

pilot_trace <- data.frame(iteration = 1:3000,
                          beta = exp(pilot$samples[, 1]),
                          gamma = exp(pilot$samples[, 2]))

ggplot(pilot_trace, aes(x = iteration, y = beta)) +
  geom_line(colour = "grey30") +
  labs(title = "Pilot run: trace of beta") +
  theme_minimal()
```

The chain finds the high-density region within the first few hundred
iterations and then moves rather sluggishly: the proposal is much wider
than the posterior, so most moves are rejected.

### Tuning the proposal from the pilot

The posterior is narrow and tilted, and the proposal should be shaped
the same way. A standard recipe (Roberts & Rosenthal, 2001) is to use
the covariance of the samples from the pilot run, scaled by $2.38^2 / d$
where $d$ is the number of parameters. This is a simple, two-stage form
of *adaptive* MCMC.

```{r}
tuned_cov <- 2.38^2 / 2 * cov(pilot$samples[-(1:1000), ])
tuned_cov
```

### Several chains from dispersed starting points

One chain cannot tell you about the places it has never visited. We run
four chains from starting points spread well beyond the posterior, a
factor of about 1.6 either side of the pilot centre, so that if they all
end up in the same place we have some evidence that it is the right
place. This is the slowest step: four chains of 6,000 iterations take a
few tens of seconds.

```{r}
centre <- colMeans(pilot$samples[-(1:1000), ])
starts <- list(centre + c(-0.5, -0.5), centre + c(0.5, -0.5),
               centre + c(-0.5, 0.5), centre + c(0.5, 0.5))

chains <- lapply(starts, function(s) {
  metropolis(log_posterior, n_iter = 6000, init = s, proposal_cov = tuned_cov)
})

sapply(chains, function(ch) ch$acceptance_rate)
```

The acceptance rates are now in the sensible range for two parameters.

### Diagnostics

First the trace plots, on the natural scale and including the burn-in so
that you can see the chains converge.

```{r}
chain_df <- do.call(rbind, lapply(seq_along(chains), function(k) {
  data.frame(chain = factor(k), iteration = 1:6000,
             beta = exp(chains[[k]]$samples[, 1]),
             gamma = exp(chains[[k]]$samples[, 2]))
}))
chain_long <- rbind(
  data.frame(chain_df[, c("chain", "iteration")], parameter = "beta", value = chain_df$beta),
  data.frame(chain_df[, c("chain", "iteration")], parameter = "gamma", value = chain_df$gamma)
)

ggplot(chain_long, aes(x = iteration, y = value, colour = chain)) +
  annotate("rect", xmin = 0, xmax = 1000, ymin = -Inf, ymax = Inf, alpha = 0.15, fill = "steelblue") +
  geom_line(alpha = 0.7) +
  facet_wrap(~ parameter, scales = "free_y", ncol = 1) +
  labs(title = "Four chains; shaded region = burn-in (first 1,000 iterations)") +
  theme_minimal()
```

The four chains start far apart and, within the burn-in, become
indistinguishable: the hairy caterpillar again, now in four colours. We
discard the first 1,000 iterations of each chain and pool the rest.

```{r}
burn_in <- 1000
posterior_samples <- do.call(rbind, lapply(chains, function(ch) ch$samples[-(1:burn_in), ]))
colnames(posterior_samples) <- c("b", "g")
nrow(posterior_samples)
```

Two numbers summarise what the plots show. The **Gelman–Rubin
statistic** $\hat{R}$ compares the variance *between* chains with the
variance *within* them; chains that have converged on the same
distribution give $\hat{R}$ close to 1, and values above about 1.01 are
a warning. The **effective sample size** allows for the fact that
consecutive MCMC draws are correlated, and so carry less information
than the same number of independent draws.

```{r}
# Gelman-Rubin R-hat for a list of chains (one parameter)
rhat <- function(chains) {
  n <- length(chains[[1]])
  W <- mean(sapply(chains, var))          # mean within-chain variance
  B <- n * var(sapply(chains, mean))      # n x variance of the chain means
  sqrt(((n - 1) / n * W + B / n) / W)
}

# Effective sample size: n over the integrated autocorrelation time
ess <- function(x) {
  rho <- acf(x, lag.max = min(length(x) - 1, 500), plot = FALSE)$acf[-1]
  k <- which(rho < 0)[1]                  # sum autocorrelations while positive
  if (!is.na(k)) rho <- rho[seq_len(k - 1)]
  length(x) / (1 + 2 * sum(rho))
}

c(rhat_b = rhat(lapply(chains, function(ch) ch$samples[-(1:burn_in), 1])),
  rhat_g = rhat(lapply(chains, function(ch) ch$samples[-(1:burn_in), 2])))

c(ess_b = ess(posterior_samples[, "b"]), ess_g = ess(posterior_samples[, "g"]))
```

We have 20,000 draws but only a few thousand *effective* ones, which is
typical of random-walk Metropolis. It is still far more than we need to
summarise two parameters.

::: callout-note
The `coda` package implements these and many other diagnostics
(`gelman.diag()`, `effectiveSize()`, `autocorr.plot()`), and `bayesplot`
draws the standard plots. We wrote our own here so that you can see
exactly what they compute.
:::

### Summarising the posterior

Everything we report is computed directly from the samples. Transform
back to the natural scale, and compute the derived quantities we care
about for every draw.

```{r}
beta_post <- exp(posterior_samples[, "b"])
gamma_post <- exp(posterior_samples[, "g"])
R0_post <- beta_post * 763 / gamma_post     # R0 = beta N / gamma
period_post <- 1 / gamma_post                # infectious period in days

summarise <- function(x) c(mean = mean(x), median = median(x),
                           quantile(x, c(0.025, 0.975)))
knitr::kable(signif(rbind(beta = summarise(beta_post),
                          gamma = summarise(gamma_post),
                          R0 = summarise(R0_post),
                          infectious_period = summarise(period_post)), 3))
```

The 2.5% and 97.5% quantiles form a **95% credible interval**, which can
be read literally: given the model, the prior and the data, there is a
95% probability that $R_0$ lies in that range. Compare this with the
maximum likelihood estimates from the previous practical.

```{r}
posterior_df <- data.frame(beta = beta_post, gamma = gamma_post, R0 = R0_post)

ggplot(posterior_df, aes(x = R0)) +
  geom_histogram(bins = 50, fill = "grey80") +
  labs(x = expression(R[0]), title = "Posterior distribution of the basic reproduction number") +
  theme_minimal()

ggplot(posterior_df, aes(x = beta, y = gamma)) +
  geom_point(alpha = 0.1, colour = "steelblue", size = 0.5) +
  labs(x = expression(beta), y = expression(gamma),
       title = sprintf("Joint posterior (correlation %.2f)", cor(beta_post, gamma_post))) +
  theme_minimal()
```

### Posterior predictive check

Finally, does the fitted model reproduce the data? For a few hundred
posterior draws we simulate the trajectory, and also the Poisson counts
around it, and plot the 95% bands against the observations.

```{r}
draws <- unname(posterior_samples[sample(nrow(posterior_samples), 500), ])

trajectories <- sapply(seq_len(nrow(draws)), function(i) {
  predict_cases(beta = exp(draws[i, 1]), gamma = exp(draws[i, 2]))
})
predictive <- apply(trajectories, 2, function(lambda) rpois(14, lambda))

bands <- data.frame(
  day = 1:14,
  traj_lower = apply(trajectories, 1, quantile, 0.025),
  traj_median = apply(trajectories, 1, quantile, 0.5),
  traj_upper = apply(trajectories, 1, quantile, 0.975),
  pred_lower = apply(predictive, 1, quantile, 0.025),
  pred_upper = apply(predictive, 1, quantile, 0.975)
)

ggplot(bands, aes(x = day)) +
  geom_ribbon(aes(ymin = pred_lower, ymax = pred_upper), fill = "tomato", alpha = 0.15) +
  geom_ribbon(aes(ymin = traj_lower, ymax = traj_upper), fill = "tomato", alpha = 0.35) +
  geom_line(aes(y = traj_median), colour = "tomato", linewidth = 1) +
  geom_point(data = flu, aes(x = day, y = cases), size = 2) +
  labs(x = "Day", y = "Pupils in bed",
       title = "Posterior fit",
       subtitle = "Dark band: 95% interval for the trajectory; light band: 95% interval for the counts") +
  theme_minimal()
```

The trajectory band is narrow, because the parameters are well
determined, and most observations fall inside the predictive band. The
peak is slightly under-predicted and the observations scatter a little
more than a Poisson model allows, which is a hint that a negative
binomial observation model might be better (see the exercises).

## Practical Considerations and Troubleshooting

**"The starting value has zero posterior density"**

- The sampler cannot start from a point where the likelihood is
  $-\infty$, for instance parameters for which the epidemic is over
  before the observed cases occur. Start closer to a plausible fit (the
  least squares or maximum likelihood estimate is ideal).

**Acceptance rate far from the target**

- Below 10%: the proposal is too wide; shrink the covariance.
- Above 70%: the proposal is too narrow; the chain is crawling. Enlarge
  it.
- Re-tune from a pilot run rather than guessing, and check the
  autocorrelation.

**Chains disagree (**$\hat{R} > 1.01$)

- Run longer, discard more burn-in, and check that no chain is stuck in
  a separate mode. If the posterior really is multimodal, a random-walk
  proposal may never cross between modes; this is a limitation of the
  method, not a bug in your code.

**Correlated parameters**

- A proposal that ignores the correlation wastes most of its moves. Use
  the pilot covariance, or update the parameters one at a time
  (*Metropolis-within-Gibbs*), or re-parameterise (e.g. in terms of
  $R_0$ and the infectious period rather than $\beta$ and $\gamma$).

**Slow likelihoods**

- Every iteration solves the ODE. Keep the pilot run short, and check
  timings before launching long chains.

## Summary

In this tutorial, we've learned how to:

1.  **Set up a Bayesian model**: a prior on the parameters and a
    likelihood for the data, and check the prior by simulating from it
2.  **Understand why the posterior is hard to compute**, and why a grid
    is not the answer
3.  **Implement the Metropolis algorithm** and see it reproduce a known
    distribution
4.  **Tune a proposal**, choose a burn-in, and diagnose convergence with
    trace plots, $\hat{R}$ and effective sample sizes
5.  **Fit the SIR model to real outbreak data by MCMC** and summarise
    the posterior: credible intervals, $R_0$, and predictions with
    uncertainty

### Key Takeaways

- **The posterior is the full answer**; estimates, intervals and
  predictions are all read off the same set of samples
- **Metropolis is a few lines of code**; the effort goes into tuning and
  diagnostics
- **Priors are assumptions**: state them, and check what they imply
- **Never trust a chain you have not checked**; diagnostics can reveal a
  problem but never prove there is none
- **Modern samplers** (adaptive Metropolis, Hamiltonian Monte Carlo and
  NUTS in Stan) automate the tuning, but do the same thing underneath

### References

- Kruschke, J. K. and Liddell, T. M. (2018). Bayesian data analysis for
  newcomers. *Psychonomic Bulletin & Review*, 25, 155–177.
- van Ravenzwaaij, D., Cassey, P. and Brown, S. D. (2018). A simple
  introduction to Markov Chain Monte–Carlo sampling. *Psychonomic
  Bulletin & Review*, 25, 143–154.
- Chib, S. and Greenberg, E. (1995). Understanding the
  Metropolis–Hastings algorithm. *The American Statistician*, 49,
  327–335.
- Roberts, G. O., Gelman, A. and Gilks, W. R. (1997). Weak convergence
  and optimal scaling of random walk Metropolis algorithms. *Annals of
  Applied Probability*, 7, 110–120.
- Roberts, G. O. and Rosenthal, J. S. (2001). Optimal scaling for
  various Metropolis–Hastings algorithms. *Statistical Science*, 16,
  351–367.
- Gelman, A. and Rubin, D. B. (1992). Inference from iterative
  simulation using multiple sequences. *Statistical Science*, 7,
  457–472.

## Exercises

- Re-run the analysis with the narrower prior on $g$ from the first
  exercise. How much do the posterior for $\gamma$ and the credible
  interval for $R_0$ change? Why so little?
- Replace the Poisson likelihood with a negative binomial one
  (`dnbinom(x, mu = predicted, size = phi, log = TRUE)`) with a third
  parameter $\phi$ (sampled on the log scale, with its own prior). Does
  the posterior predictive band cover the data better? What happens to
  the credible intervals for $\beta$ and $\gamma$?
- Run the four chains with the untuned pilot proposal
  `diag(c(0.1, 0.1)^2)` instead of `tuned_cov`. Compare the acceptance
  rates, the trace plots and the effective sample sizes.
- Thin the pooled sample by keeping every 10th draw. How much does the
  effective sample size fall compared with the number of draws you threw
  away?
- Write a *Metropolis-within-Gibbs* version of the sampler that proposes
  a move for $b$ and $g$ one at a time, each with its own proposal
  standard deviation. Does it mix better or worse than the joint
  proposal?
- Fit an SEIR model with a fixed latent period of one day. Compare the
  posterior for $R_0$ with the SIR result.

## Acknowledgements

The structure of this tutorial follows the [least squares and maximum
likelihood practical](01_least_squares_mle_practicals.qmd), which is based
on
[materials](https://ms.mcmaster.ca/~bolker/eeid/2011_eco/EEID2011_Fitting.pdf)
by [John M. Drake](https://daphnia.ecology.uga.edu/drakelab/). The
explanations draw on the three papers listed above and on the course
[Model fitting and inference for infectious disease
dynamics](https://mfiidd.github.io/mfiidd/) by Sebastian Funk and
colleagues.
